Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-06-28 manuscript/main.tex
V^\ast to \mathbf{V}^\ast

	Furthermore, let $\widetilde{Z} = [\mathbf{Z}_1, \cdots, \mathbf{Z}_n] \in \mathbb{R}^{n \times p}$ denote $n \times p$ matrix of the scalar predictors (observed version when $l=1$, or residualized version when $l \ge 2$), and let $\widetilde{W}[\mathbf{t}]=[\widetilde{X}(\mathbf{t}), \widetilde{Z}]$ denote $n \times (K+p)$ hybrid predictor matrix that stacks $\widetilde{X}(\mathbf{t})$ and $\widetilde{Z}$ as columns. Then we can also write the hybrid predictor, whose functional part is approximated using the basis functions, as a form of linear transformation similar to \eqref{eq: basis approximation of X}:
	\begin{equation}\label{eq: basis approximation of W}
		\widetilde{W}[\mathbf{t}] =  \widetilde{C}^* B^*[\mathbf{t}], \quad t \in \mathcal{T}.
	\end{equation}
	where  $B^*[\mathbf{t}] = \mathrm{blkdiag}(B(\mathbf{t}), I_p) \in \mathbb{R}^{(MK+p) \times (K+p)}$ and $\widetilde{C}^*=[\widetilde{C}, \widetilde{Z}] \in \mathbb{R}^{n \times (MK+p)}$.

@@

We now describe the strategy for computing the PLS direction at the $l$-th iteration. For simplicity, we omit the iteration index and denote the residualized responses by $\mathbf{Y} \in \mathbb{R}^n$, 
and
let $\widetilde{\mathbf{C}}^\ast$ denote
the coefficients of the basis approximation of the residualized hybrid predictors by . 
 Our strategy finds the direction $\boldsymbol{\xi} \in \mathcal{H}$ that minimizes the empirical covariance between the 
\begin{align*}
	\widehat{\textnormal{Cov}} (\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, \mathbf{Y} ) &= \frac{1}{n} \sum \limits_{i=1}^n \langle \mathbf{W}_i, \boldsymbol{\xi} \rangle_\mathcal{H} Y_i
\end{align*}

@@

		\jongmin{in the (1,1)th block, should $ \mathbf{y}$ be replaced by $\widetilde{\mathbf{y}} $? }
		\begin{equation}\label{def:V_star}
			V^\ast = n^{-2}\begin{bmatrix}
				J \widetilde{C}^\top \mathbf{y}\mathbf{y}^\top \widetilde{C} J & J \widetilde{C}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z} \\
				\widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C} J &   \widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z}
			\end{bmatrix} = n^{-2} J^\ast \widetilde{C}^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C}^{*} J^\ast
			\in \mathbb{R}^{(MK+p) \times (MK+p)}.
		\end{equation}
		and  $\boldsymbol{\xi}^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th PLS component, $\widehat{\boldsymbol{\xi}} \equiv \widehat{\boldsymbol{\xi}}_l$ can be obtained by solving
		\begin{equation}\label{def:eigenproblem_smoothed}
			\argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \; \frac{\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*}{\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*}, \quad \textit{or equivalently}, \quad \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \;  \boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^* \quad \textnormal{subject to}  \quad \boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1,
		\end{equation}
		with respect to $\widehat{\boldsymbol{\xi}}^*=(\widehat{\mathbf{a}}^{\top}, \widehat{\boldsymbol{\theta}}^{\top})^\top$, and then setting $\widehat{\boldsymbol{\xi}}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^*[\mathbf{t}]^\top \widehat{\boldsymbol{\xi}}^*$. The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as $\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}$.
	\end{proposition}
	The proof of Proposition \ref{Proposition: generalized rayleigh quotient} is given in the Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.

>>>>

	Furthermore, let $\widetilde{Z} = [\mathbf{Z}_1, \cdots, \mathbf{Z}_n] \in \mathbb{R}^{n \times p}$ denote $n \times p$ matrix of the scalar predictors (observed version when $l=1$, or residualized version when $l \ge 2$), and let $\widetilde{W}[\mathbf{t}]=[\widetilde{X}(\mathbf{t}), \widetilde{Z}]$ denote $n \times (K+p)$ hybrid predictor matrix that stacks $\widetilde{X}(\mathbf{t})$ and $\widetilde{Z}$ as columns. Then we can also write the hybrid predictor, whose functional part is approximated using the basis functions, as a form of linear transformation similar to \eqref{eq: basis approximation of X}:
	\begin{equation}\label{eq: basis approximation of W}
		\widetilde{W}[\mathbf{t}] 
		=  
		\widetilde{ \mathbf{C} }^\ast B^*[\mathbf{t}], \quad t \in \mathcal{T}.
	\end{equation}
	where  $B^*[\mathbf{t}] = \mathrm{blkdiag}(B(\mathbf{t}), I_p) \in \mathbb{R}^{(MK+p) \times (K+p)}$ and $\widetilde{ \mathbf{C} }^\ast=[\widetilde{C}, \widetilde{Z}] \in \mathbb{R}^{n \times (MK+p)}$.

@@

We now describe the strategy for computing the PLS direction at the $l$-th iteration. For simplicity, we omit the iteration index and denote the residualized response by $Y$, with its finite-sample observations denoted by $\mathbf{y} = (y_1, \ldots, y_n) \in \mathbb{R}^n$. Let $\widetilde{W} \in \mathcal{H}$ denote the residualized, basis-approximated hybrid predictor, and let $\widetilde{\mathbf{C}}^\ast \in \mathbb{R}^{n \times (MK+p)}$ denote the basis coefficients of its finite-sample observations. Our strategy finds the direction $\boldsymbol{\xi} \in \mathcal{H}$ that maximizes the empirical covariance between the resulting PLS component $\langle \widetilde{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response.

For practical computation, we first construct the following matrix:
	\begin{equation}\label{def:V_star}
	V^\ast = n^{-2}\begin{bmatrix}
		J \widetilde{C}^\top \mathbf{y}\mathbf{y}^\top \widetilde{C} J & J \widetilde{C}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z} \\
		\widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C} J &   \widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z}
	\end{bmatrix} = n^{-2} J^\ast \widetilde{C}^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C}^{*} J^\ast
	\in \mathbb{R}^{(MK+p) \times (MK+p)}.
\end{equation}

@@

		\jongmin{in the (1,1)th block, should $ \mathbf{y}$ be replaced by $\widetilde{\mathbf{y}} $? }

		and  $\boldsymbol{\xi}^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th PLS component, $\widehat{\boldsymbol{\xi}} \equiv \widehat{\boldsymbol{\xi}}_l$ can be obtained by solving
		\begin{equation}\label{def:eigenproblem_smoothed}
			\argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \; \frac{\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*}{\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*}, \quad \textit{or equivalently}, \quad \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \;  \boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^* \quad \textnormal{subject to}  \quad \boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1,
		\end{equation}
		with respect to $\widehat{\boldsymbol{\xi}}^*=(\widehat{\mathbf{a}}^{\top}, \widehat{\boldsymbol{\theta}}^{\top})^\top$, and then setting $\widehat{\boldsymbol{\xi}}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^*[\mathbf{t}]^\top \widehat{\boldsymbol{\xi}}^*$. The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as $\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}$.
	\end{proposition}
	The proof of Proposition \ref{Proposition: generalized rayleigh quotient} is given in the Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.
